---
title: 'Assignment 2A: SQL & R'
author: "Ummay Rukiya"
date: "2026-09-09"
output: html_document
---

## Overview

For this assignment, I plan to collect movie ratings from at least five people and store the results in a SQLite database. The six selected movies are *The Odyssey*, *Project Hail Mary*, *Toy Story 5*, *Spider-Man: Brand New Day*, *The Super Mario Galaxy Movie*, and *Moana*. Each participant will rate every movie they have seen on a scale from 1 to 5.

## Planned Approach

I will ask at least five people which of the six movies they have seen and collect their ratings. To protect their privacy, I will identify them as Participant 1, Participant 2, and so on instead of using their names.

I plan to use SQLite because it is supported for this assignment and can be used directly through R. I will create three related tables: a users table, a movies table, and a ratings table. The users and movies tables will contain unique identification numbers, while the ratings table will connect each participant to each movie and store the rating.

I will include the SQL commands used to create and populate all three tables. After storing the data, I will use R with the DBI and RSQLite packages to connect to the database, join the tables, and load the results into a dataframe. I will then inspect the data and calculate simple summaries, including the number of ratings and average rating for each movie.

## Missing-Data Strategy

Some participants may not have seen every movie. These missing ratings will be stored as `NULL` in SQL and will appear as `NA` after the data is loaded into R. I will keep track of the missing values but exclude them when calculating average movie ratings so that unseen movies are not treated as low ratings.

## Anticipated Challenges

The main challenges may include configuring the SQLite connection, correctly connecting the three tables through identification numbers, and making sure all ratings are either integers from 1 to 5 or missing. I will also need to make sure that the SQL code can recreate the tables and that no passwords or private participant information appear in the submitted files.

## Database Setup

I used SQLite because it does not require a password or a separate database server. The database is created in memory whenever this document runs.

```{r database-setup}
library(DBI)
library(RSQLite)

con <- dbConnect(RSQLite::SQLite(), ":memory:")
dbExecute(con, "PRAGMA foreign_keys = ON;")
```

## Creating the SQL Tables

I created separate tables for participants, movies, and ratings. The ratings table connects each participant to each movie using identification numbers.

```{r create-tables}
dbExecute(con, "DROP TABLE IF EXISTS ratings;")
dbExecute(con, "DROP TABLE IF EXISTS participants;")
dbExecute(con, "DROP TABLE IF EXISTS movies;")

dbExecute(con, "
CREATE TABLE participants (
  participant_id INTEGER PRIMARY KEY,
  participant_label TEXT NOT NULL UNIQUE
);
")

dbExecute(con, "
CREATE TABLE movies (
  movie_id INTEGER PRIMARY KEY,
  movie_title TEXT NOT NULL UNIQUE
);
")

dbExecute(con, "
CREATE TABLE ratings (
  participant_id INTEGER NOT NULL,
  movie_id INTEGER NOT NULL,
  rating REAL,
  PRIMARY KEY (participant_id, movie_id),
  CHECK (rating IS NULL OR rating BETWEEN 1 AND 5),
  FOREIGN KEY (participant_id) REFERENCES participants(participant_id),
  FOREIGN KEY (movie_id) REFERENCES movies(movie_id)
);
")
```

## Adding Participants and Movies

I used participant labels instead of personal names. Each movie and participant has a unique identification number.

```{r add-participants-and-movies}
dbExecute(con, "
INSERT INTO participants (participant_id, participant_label)
VALUES
  (1, 'Participant 1'),
  (2, 'Participant 2'),
  (3, 'Participant 3'),
  (4, 'Participant 4'),
  (5, 'Participant 5');
")

dbExecute(con, "
INSERT INTO movies (movie_id, movie_title)
VALUES
  (1, 'The Odyssey'),
  (2, 'Project Hail Mary'),
  (3, 'Toy Story 5'),
  (4, 'Spider-Man: Brand New Day'),
  (5, 'The Super Mario Galaxy Movie'),
  (6, 'Moana');
")
```

## Adding the Movie Ratings

The collected ratings are stored exactly as provided, including decimal ratings. A missing rating is stored as SQL `NULL` when a participant has not seen a movie.

```{r add-ratings}
dbExecute(con, "DELETE FROM ratings;")

dbExecute(con, "
INSERT INTO ratings (participant_id, movie_id, rating)
VALUES
  (1, 1, 4.5),
  (2, 1, 4.9),
  (3, 1, 3.5),
  (4, 1, 5.0),
  (5, 1, 4.5),

  (1, 2, 5.0),
  (2, 2, 5.0),
  (3, 2, 4.5),
  (4, 2, 4.8),
  (5, 2, 3.5),

  (1, 3, 4.0),
  (2, 3, NULL),
  (3, 3, 3.5),
  (4, 3, 4.5),
  (5, 3, 3.5),

  (1, 4, 5.0),
  (2, 4, NULL),
  (3, 4, 4.8),
  (4, 4, 4.5),
  (5, 4, 5.0),

  (1, 5, 3.5),
  (2, 5, 3.0),
  (3, 5, 3.5),
  (4, 5, NULL),
  (5, 5, 3.0),

  (1, 6, 2.5),
  (2, 6, 3.0),
  (3, 6, 3.3),
  (4, 6, NULL),
  (5, 6, NULL);
")
```

## Loading the SQL Data into R

I used a SQL query with joins to combine the three tables. The result is loaded into R as a dataframe named `movie_ratings`.

```{r load-data}
movie_ratings <- dbGetQuery(con, "
SELECT
  p.participant_id,
  p.participant_label,
  m.movie_id,
  m.movie_title,
  r.rating
FROM ratings AS r
JOIN participants AS p
  ON r.participant_id = p.participant_id
JOIN movies AS m
  ON r.movie_id = m.movie_id
ORDER BY p.participant_id, m.movie_id;
")

head(movie_ratings, 12)
```

## Handling Missing Ratings

Ratings marked as “Not Seen” were stored as SQL `NULL`. After the data was loaded into R, these values became `NA`. I kept them as missing rather than replacing them with zero because not seeing a movie is different from giving it a low rating.

```{r missing-ratings}
movie_ratings$rating_status <- ifelse(
  is.na(movie_ratings$rating),
  "Not Seen",
  "Rated"
)

missing_ratings <- movie_ratings[
  is.na(movie_ratings$rating),
  c("participant_label", "movie_title", "rating_status")
]

missing_ratings
```

## Movie Rating Summary

For each movie, I calculated how many ratings were received, how many ratings were missing, and the average of the available ratings. Missing ratings were removed only when calculating the averages.

```{r movie-summary}
movie_summary <- do.call(
  rbind,
  lapply(split(movie_ratings, movie_ratings$movie_title), function(movie) {
    data.frame(
      movie_title = movie$movie_title[1],
      ratings_received = sum(!is.na(movie$rating)),
      missing_ratings = sum(is.na(movie$rating)),
      average_rating = round(mean(movie$rating, na.rm = TRUE), 2)
    )
  })
)

row.names(movie_summary) <- NULL
movie_summary <- movie_summary[
  order(movie_summary$average_rating, decreasing = TRUE),
]

movie_summary
```

## Conclusions and Next Steps

I successfully collected movie ratings from five participants, stored them in three related SQLite tables, and loaded the combined data into an R dataframe. Out of 30 possible ratings, 25 movies were rated and five were marked as not seen. Missing ratings were kept as `NULL` in SQL and `NA` in R so they would not incorrectly lower the averages.

*Spider-Man: Brand New Day* received the highest average rating at 4.82, while *Moana* received the lowest average rating at 2.93. These results only represent a small group of five people, so they cannot represent the opinions of a larger population. I could extend this project by collecting ratings from more people and comparing whether participants have different rating habits.

```{r close-connection}
dbDisconnect(con)
```








